EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-08-20 jlamp/jlamp.tex
minima

\noindent
\begin{minipage}[l]{.85\linewidth}
\begin{itemize}
	\item		\emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one aside, if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
	\item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable;
	\item \emph{closed under decomposition}, or \emph{left-cancellable}, if whenever $g$ and $g\circ f$ belong to $\mathcal{A}$, then $f\in \mathcal{A}$.
	\end{itemize}
\end{minipage}\hfill 
\begin{minipage}[r]{.13\linewidth}
\xymatrix{A \ar[r]^{f} \ar[d]_{m}& B \ar[d]^{n} \\ C \ar[r]_{g} & D}
\end{minipage}
%\begin{itemize}
%\end{itemize}

	\begin{defi}
	Let $\mathcal{A}\subseteq \mor(\X)$ be a class of arrows in a category $\X$ and consider the cube below. %on the right. 
	We say that the bottom square is an \emph{$\mathcal{A}$-Van Kampen square} if
\vspace{-.5cm}

	\begin{minipage}[l]{.68\linewidth}
	\begin{enumerate}
		\item it is a pushout square;
		\item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
		if and only if the front and right faces are pullbacks.
	\end{enumerate}
\end{minipage}\hfill 
\begin{minipage}[r]{.3\linewidth}
	\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }
\end{minipage}
	Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}. A $\mor(\X)$-Van Kampen square is called  \emph{Van
		Kampen} and a $\mor(\X)$-stable square  \emph{stable}.
\end{defi}

@@

 \smallskip 
 We can interpret an arbitrary hypergaph as a \emph{typed} or \emph{colored} monoidal signature: the set of nodes provides the \emph{types} or \emph{colors}, while hyperedges can be thought of as operations, with arity and coarity provided by the source and target functions, respectively. This suggests the following definition, which will be used in Section \ref{par:tg}\label{sub:hyp_alg}.

@@

\paragraph{Separated hypergraphs}
We can now introduce the second  subcategory of $\hyp$ in which we are interested in and that will be used in \Cref{sec:sepegg} to give us the underlying hypergraphs for EGGs with minimally identified tokens.

\begin{defi}  A left-monogamous hypergraph $\mathcal{G}=(E_{\mathcal{G}}, V_{\mathcal{G}}, s_{\mathcal{G}}, t_{\mathcal{G}})$ is \emph{separated} if
	\begin{enumerate}
	\item it is \emph{right-monogamous}, i.e.~$(E_{\mathcal{G}}, V_{\mathcal{G}}, t_{\mathcal{G}}, s_{\mathcal{G}})$ is left-monogamous;
	\item for every pair of, not necessarily distinct, hyperedges $e$ and $e'$,  $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e')$ are disjoint.
	\end{enumerate}
We will denote by $\shyp$ the full subcategory of $\hyp$ given by separated hypergraphs and by $S$ the inclusion functor.
\end{defi}

@@

 \begin{rem}\label{rem:elem2}
 	Let $\mathcal{G}$ be a separated hypergraph,  
 	applying \Cref{rem:elem} we this is equivalent to say that, for every hyperedge $e$:
 	\begin{itemize}
 		\item $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ do not contain repeated letters;
 		\item $s_{\mathcal{G}}(e)$ is disjoint from $t_{\mathcal{G}}(e')$ for every $e'\in E_{\mathcal{G}}$;
 		\item for every $e'\in E_{\mathcal{G}}$ different from $e$, $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ are disjoint from, respectively $s_{\mathcal{G}}(e')$ and $t_{\mathcal{G}}(e')$.
 	\end{itemize}
 \end{rem}

@@

\begin{defi}
Given an hypergraph $\mathcal{\mathcal{G}}=(E_{\mathcal{G}}, V_{\mathcal{F}}, s_{\mathcal{G}}, t_{\mathcal{G}})$, for every hyperedge $e\in E_{\mathcal{G}}$ with $n_e=\lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ and $m_e=\lgh_{V_{\mathcal{G}}}(t_{\mathcal{G}}(e))$, we can consider the coproduct $(V_e, \{\iota^e_{i}\}_{i=1}^{n_e+m_e})$ of $n_e+m_e$ copies of $1$.
We can further take the coproduct $(V_{\sp(\mathcal{G})}, \{\iota_e\}_{e\in E_{\mathcal{G}}})$ of the family $\{V_{e}\}_{e\in E_\mathcal{G}}$ and define:
\[\begin{split}
	s_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=1}^{n_e} \iota_e(\iota^e_{i}(0))
\end{split}\qquad \begin{split}
	t_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=n_e+1}^{n_e + m_e} \iota_e(\iota^e_{i}(0))
\end{split} \]
where $0$ is the unique element of $1$.

We define the \emph{separated graph associated to $\mathcal{G}$}  as $\sp(\mathcal{G}):= (E_{\mathcal{G}}, V_{\sp(\mathcal{G})}, s_{\sp(\mathcal{G})}, t_{\sp(\mathcal{G})})$.
\end{defi}
\begin{rem}\label{rem:ob}
 By construction $\sp(\mathcal{G})$ is an object of $\shyp$. To see this it is enough to apply \Cref{rem:elem2}:
 \begin{itemize}
 	\item every letter of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ lies in the image of a different coprojection $\iota_e\circ \iota^e_i$, so it cannot appear more than one time;
 	\item a letter of $s_{\mathcal{G}}(e)$ are all contained in the image of $\iota_{e}\circ \iota^e_i$  for $i\leq n_e$, while every letter of $t_{\mathcal{G}}(e)$ lies in the image of $\iota_{e}\circ \iota^e_j$  for $n_e < j$, guaranteeing disjointness;
 	\item if $e$ is different from $e'$, then all the letters of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ are contained in the image of $\iota_e$, while those of  $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ are in the image of $\iota_{e'}$ and so, again, we get the wanted disjointness. 
 \end{itemize}
\end{rem}

@@

	We can then depict $\mathcal{G}^\Sigma $ and $\sp(\mathcal{G}^\Sigma )$ as below.
	\begin{center}	\begin{tikzpicture}
			\begin{pgfonlayer}{nodelayer}
				\node[style=node] (v) at (0, 0){};
				\node[style=none] (vlab) at (0, -0.4){$v$};
				\node[style=small box] (n) at (-1.5, 0) {};
				\node[style=none] (nlab) at (-1.5, 0.5) {$e$};
				\node[style=none] (nattach) at (-1.6, 0) {};

@@

		Let $\mathcal{H}$ be $(\emptyset, 1, ?_1, ?_1)$, which is clearly a separated hypergraph. Now,  $\shyp(\mathcal{H}, \sp(\mathcal{D}))$ has the same cardinality of  $\shyp(\mathcal{H}, \mathcal{G})$, which is $4$, while $\hyp(S(\mathcal{H}), \mathcal{D})$ is a singleton.
		\item Consider now the ``single loop'' $\mathcal{L}=(1, 1, \id{1}, \id{1})$. Then $\sp(\mathcal{L})$ is isomorphic to $(1, 2, \id{2}, \id{2})$ and so there are no arrows $\mathcal{L}\to S(\sp(\mathcal{L}))$. In particular this shows that $\sp$ cannot be left adjoint to $S$. 
	\end{itemize} 
\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.27\linewidth}
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node[style=node] (v) at (0, 0){};
			\node[style=none] (vlab) at (0, 0.3){$0$};
			\node[style=small box] (times) at (-1, 1) {};
			\node[style=none] (timeslab) at (-1, 1.5) {$0$};
			\node[style=none] (timesin) at (-1, .7) {};
			\node[style=none] (timesout) at (-.7, 1){};
			\node[style=small box] (div) at (1, 1) {};
			\node[style=none] (divlab) at (1, 1.5) {$1$};
			\node[style=none] (divin) at (1, .7) {};
			\node[style=none] (divout) at (.7, 1){};

@@

\smallskip
If we restrict our attention to separated hypergraph without isolated nodes, then $\sp$ becomes a right adjoint. The interested reader can consult \Cref{proof:dx} for a proof.

\begin{defi} Let $\mathcal{G}$ be an hypergraph. A node $v\in V_{\mathcal{G}}$ is \emph{isolated} if for every $e\in E_{\mathcal{G}}$ $v$ does not occur as a letter in neither $s_{\mathcal{G}}$ nor in $t_{\mathcal{G}}$. $\mathcal{G}$ is \emph{pruned} if it does not have isolated nodes. We denote by $\phyp$ and $\pshyp$ the full subcategories of $\hyp$ given by, respectively, pruned hypergraphs and those which are pruned and separated.
\end{defi}

\begin{rem}\label{rem:prun}
	By construction, $\sp{G}$ is pruned for every hypergraph $\mathcal{G}$.
\end{rem}

@@

Finally, we conclude our exam of separated hypergraph proving a result regarding connected limits. A proof of it is \Cref{proof:seplim}

@@

Since $\mathcal{T}$ must be separated, we have two (non disjoint) cases:

\begin{minipage}[l]{.65\linewidth}
	\begin{itemize}
		\item  all the hyperedges of $\mathcal{T}$ have sources with length zero; 
		\item all the hyperedges of $\mathcal{T}$ have targets with length zero.
	\end{itemize}
\end{minipage}
\begin{minipage}[r]{.15\linewidth}
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node[style=medium box] (div) at (5, 1.3) {};
		\node[style=none] (divlab) at (5, 2) {$e$};
		\node[style=none] (divin1) at (4.7, 1.6) {};
		\node[style=none] (divin2) at (4.7, 1) {};
		\node[style=none] (divout) at (5.2, 1.3){};

@@

\smallskip 
\noindent 
\begin{minipage}[l]{.58\linewidth}
\begin{exa}\label{ex:nopush}
	In general $\shyp$ does not have pushout. For instance consider the solid part of diagram aside. Let us call the two unnamed hypergraphs on the top row $\mathcal{H}$ and $\mathcal{K}$, respectively, while will call the third one $\mathcal{Q}$.

	Suppose that there exists $\mathcal{G}$ and arrows $(p_1, p_2)\colon \mathcal{Q}\to \mathcal{G}$ and $(q_1, q_2)\colon \mathcal{K}\to \mathcal{G}$ which make the square a pushout in $\shyp$. 
	By construction we have
	\[s_{\mathcal{G}}(q_1(e)) = q^\star_2(s_{\mathcal{K}}(e)) = q^\star_2(j^\star_2(v))=p^\star_2(i^\star_2(v)) = p^\star_2(t^\star_{\mathcal{Q}}(e))= t_{\mathcal{G}}(p_2(e))\]
	and so $\mathcal{G}$ cannot ber in $\shyp$, yielding a contradiction.
\end{exa}
\end{minipage} \hfill \begin{minipage}[r]{.5\linewidth} \vspace{-2cm}
\xymatrix@R=6pt@C=30pt{\begin{tikzpicture}
		\node[style=node](v) at (0, 0){};
	\node[style=none] at(0, 0.3){$v$};
	\end{tikzpicture}\ar[d]_{(i_1, i_2)} \ar[r]^-{(j_1, j_2)}  &
	\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$j_2(v)$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$w$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, 0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw[style=none] (v) to (in.east); 
		\draw[style=none] (out.east) to (u.west);
	\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[d]^{(q_1, q_2)} \\\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$u$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$i_2(v)$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, -0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw[style=none] (v) to (in.east); 
			\draw[style=none] (out.east) to (u.west);
		\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[r]_-{(p_1, p_2)}&
	\mathcal{G}
	}
\end{minipage}

@@

Our interest for hypergraphs stems from their use as a graphical representation of algebraic terms. As  noted in \Cref{sub:hyp_alg}, an hypergraph can be thought as a colored monoidal signature. Thus labelling correspnds to the operationn of taking the slice category over a suitable hypergraph.

@@

\Cref{cor:mono} and \Cref{thm:slice-functors} give us immediately an adhesivity result for $\hyp_{\mathcal{G}}$ and a characterisation of monos in it.

@@

			Consider an algebraic signature $\Sigma$ and an hypergraph $\mathcal{H}=(E, V, s, t)$. By definition we know that $U_{\hyp}(\mathcal{G}^{\Sigma})$ is the terminal object $1$, so an arrow $\mathcal{H}\rightarrow \mathcal{G}^{\Sigma}$, is determined by a function $h\colon E_\mathcal{H}\to O_\Sigma$  making the two squares on the right commutative (cfr.~\Cref{rem:length}).

@@

As we have done in \Cref{sub:sub}, we can restrict our attention to two subcategories of labelled hypergraphs.

@@

\begin{prop}
Let $M\colon \lmo \to \hyp$ and $S\colon \shyp \to \hyp$ be the inclusion functors. Let, moreover, $\Delta_{\mathcal{G}}\colon 1\to \hyp$ be the constant functor in $\mathcal{G}$. Then:
\begin{enumerate}
	\item $\lmo_{\mathcal{G}}$ is equivalent to $\comma{M}{\Delta_\mathcal{G}}$;
	\item $\shyp_{\mathcal{G}}$ is equivalent to $\comma{S}{\Delta_\mathcal{G}}$; 
\end{enumerate}
\end{prop}

\Cref{lim} allows us to easily characterize regular monos in $\hyp_{\mathcal{G}}$.
\begin{prop}
	A morphism $(f,g)$ between $(h_1, k_1)\colon \mathcal{H}\to \mathcal{G}$ and  $(h_2, k_2)\colon \mathcal{K}\to \mathcal{G}$ is a regular mono in $\lmo_{\mathcal{G}}$ if and only if $(f,g)\colon \mathcal{H}\to \mathcal{K}$ is a regular mono in $\lmo$.
\end{prop}

In turn, by \Cref{thm:slice-functors}, Corollaries \ref{cor:eq}, \ref{cor:pusho} and \Cref{lem:regmon} we get the following.

\begin{thm} Given an hypergraph $\mathcal{G}$, the following hold true:
	\begin{enumerate}
		\item an arrow $(f,g)$  if and only if is so in $\lmo$.
		\item $\lmo_{\mathcal{G}}$ is quasiadhesive.
	\end{enumerate}
\end{thm}

\begin{rem}\label{rem:tg}
	Let $\Sigma$ be an algebraic signature. Then we know that every hypergraph labelled by $\Sigma$ must be algebraic. By \Cref{rem:elem} this means that an hypergraph $(h,k)\colon \mathcal{H}\to \mathcal{G}^\Sigma$ labelled by $\Sigma$ is left-monogamous if and only if $t_{\mathcal{H}}$ is injective. This is precisely the definition of (cyclic) \emph{term graphs} in \cite{Plu:TGR-ENTCS,CorradiniG05,CastelnovoGM24}. 
\end{rem}

@@

In turn,this allows us to define $\epsilon_{G}\colon V_{\sp(\mathcal{G})}\to V_{\mathcal{G}}$ as the unique one fitting in the diagram below.

>>>>

\smallskip 
\noindent
\begin{minipage}[l]{.85\linewidth}
\begin{itemize}
	\item		\emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one aside, if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
	\item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable;
	\item \emph{closed under decomposition}, or \emph{left-cancellable}, if whenever $g$ and $g\circ f$ belong to $\mathcal{A}$, then $f\in \mathcal{A}$.
	\end{itemize}
\end{minipage}\hfill 
\begin{minipage}[r]{.13\linewidth}
\xymatrix{A \ar[r]^{f} \ar[d]_{m}& B \ar[d]^{n} \\ C \ar[r]_{g} & D}
\end{minipage}
%\begin{itemize}
%\end{itemize}

\medskip
\begin{minipage}[l]{.73\linewidth}
	\begin{defi}
	Let $\mathcal{A}\subseteq \mor(\X)$ be a class of arrows in a category $\X$ and consider the cube aside. 
	%We say that the 
	The bottom square is an \emph{$\mathcal{A}$-Van Kampen square} if
%\vspace{-.5cm}	
	\begin{enumerate}
		\item it is a pushout square;
		\item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
		if and only if the front and right faces are pullbacks.
	\end{enumerate}
\end{defi}
\end{minipage}
\hfill 
\begin{minipage}[r]{.25\linewidth}
\vspace{-.cm}
	\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }
\end{minipage}

\medskip
Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}. A $\mor(\X)$-Van Kampen square is called  \emph{Van
Kampen} and a $\mor(\X)$-stable square  \emph{stable}.

@@

 \smallskip 
 We can interpret an arbitrary hypergaph as a \emph{typed} or \emph{colored} monoidal signature: the nodes provide the \emph{types} or \emph{colors}, while hyperedges can be thought of as operations, with arity and coarity provided by the source and target functions, respectively. This suggests the following definition, which is used later on.
 % in Section \ref{par:tg}\label{sub:hyp_alg}.

@@

\paragraph{Separated hypergraphs}
We can now introduce the second  subcategory of $\hyp$ in which we are interested in and that is going to be used in \Cref{sec:sepegg} to give the underlying hypergraphs for EGGs with minimally identified tokens.

\begin{defi}  A left-monogamous hypergraph $\mathcal{G}=(E_{\mathcal{G}}, V_{\mathcal{G}}, s_{\mathcal{G}}, t_{\mathcal{G}})$ is \emph{separated} if
	\begin{enumerate}
	\item it is \emph{right-monogamous}, i.e.~$(E_{\mathcal{G}}, V_{\mathcal{G}}, t_{\mathcal{G}}, s_{\mathcal{G}})$ is left-monogamous;
	\item for every pair of, not necessarily distinct, hyperedges $e$ and $e'$,  $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e')$ are disjoint.
	\end{enumerate}
We denote by $\shyp$ the full subcategory of $\hyp$ of separated hypergraphs and by $S$ the inclusion functor.
\end{defi}

@@

 \begin{rem}\label{rem:elem2}
 	Let $\mathcal{G}$ be a separated hypergraph.
 	By applying \Cref{rem:elem}, this fact is equivalent to say that for every hyperedge $e$
 	\begin{itemize}
 		\item $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ do not contain repeated letters;
 		\item $s_{\mathcal{G}}(e)$ is disjoint from $t_{\mathcal{G}}(e')$ for every $e'\in E_{\mathcal{G}}$;
 		\item for every $e'\in E_{\mathcal{G}}$ different from $e$, $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ are disjoint from, respectively $s_{\mathcal{G}}(e')$ and $t_{\mathcal{G}}(e')$.
 	\end{itemize}
 \end{rem}

@@

\begin{defi}
Given an hypergraph $\mathcal{\mathcal{G}}=(E_{\mathcal{G}}, V_{\mathcal{F}}, s_{\mathcal{G}}, t_{\mathcal{G}})$, for every hyperedge $e\in E_{\mathcal{G}}$ with $n_e=\lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ and $m_e=\lgh_{V_{\mathcal{G}}}(t_{\mathcal{G}}(e))$, we can consider the coproduct $(V_e, \{\iota^e_{i}\}_{i=1}^{n_e+m_e})$ of $n_e+m_e$ copies of $1$.
We can further take the coproduct $(V_{\sp(\mathcal{G})}, \{\iota_e\}_{e\in E_{\mathcal{G}}})$ of the family $\{V_{e}\}_{e\in E_\mathcal{G}}$ and define
\[\begin{split}
	s_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=1}^{n_e} \iota_e(\iota^e_{i}(0))
\end{split}\qquad \begin{split}
	t_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=n_e+1}^{n_e + m_e} \iota_e(\iota^e_{i}(0))
\end{split} \]
where $0$ is the unique element of $1$. 
\marginpar{ovvero?}

We define the \emph{separated graph associated to $\mathcal{G}$}  as $\sp(\mathcal{G}) = (E_{\mathcal{G}}, V_{\sp(\mathcal{G})}, s_{\sp(\mathcal{G})}, t_{\sp(\mathcal{G})})$.
\end{defi}
\begin{rem}\label{rem:ob}
 By construction $\sp(\mathcal{G})$ is an object of $\shyp$. To see this it is enough to apply \Cref{rem:elem2}
 \begin{itemize}
 	\item every letter of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ lies in the image of a different coprojection $\iota_e\circ \iota^e_i$, so it cannot appear more than one time;
 	\item every letter of $s_{\mathcal{G}}(e)$ lies in the image of $\iota_{e}\circ \iota^e_i$  for $i\leq n_e$, while every letter of $t_{\mathcal{G}}(e)$ lies in the image of $\iota_{e}\circ \iota^e_j$  for $n_e < j$, guaranteeing disjointness;
 	\item if $e$ is different from $e'$, then all the letters of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ are  in the image of $\iota_e$, while those of  $s_{\mathcal{G}}(e')$ and $t_{\mathcal{G}}(e')$ are in the image of $\iota_{e'}$ and so, again, we get the required disjointness. 
 \end{itemize}
\end{rem}

@@

	We can then depict $\mathcal{G}^\Sigma $ and $\sp(\mathcal{G}^\Sigma )$ as below\marginpar{e le lettere?}
	\begin{center}	\begin{tikzpicture}
			\begin{pgfonlayer}{nodelayer}
				\node[style=node] (v) at (0, 0){};
				\node[style=none] (vlab) at (0, -0.4){$v$};
				\node[style=small box] (n) at (-1.5, 0) {};
				\node[style=none] (nlab) at (-1.5, 0.5) {$e$};
				\node[style=none] (nattach) at (-1.6, 0) {};

@@

		Let $\mathcal{H}$ be $(\emptyset, 1, ?_1, ?_1)$, which is clearly a separated hypergraph. Now,  $\shyp(\mathcal{H}, \sp(\mathcal{D}))$ has the same cardinality of  $\shyp(\mathcal{H}, \mathcal{G})$, which is $4$, while $\hyp(S(\mathcal{H}), \mathcal{D})$ is a singleton.
		\item Consider now the ``single loop'' $\mathcal{L}=(1, 1, \id{1}, \id{1})$. Then $\sp(\mathcal{L})$ is isomorphic to $(1, 2, \id{2}, \id{2})$ and so there are no arrows $\mathcal{L}\to S(\sp(\mathcal{L}))$. In particular, this shows that $\sp$ cannot be left adjoint to $S$. 
	\end{itemize} 
\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.27\linewidth}
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node[style=node] (v) at (0, 0){};
			\node[style=none] (vlab) at (0, 0.3){$0$};
			\node[style=small box] (times) at (-1, 1) {};
			\node[style=none] (timeslab) at (-1, 1.5) {$0$};
			\node[style=none] (timesin) at (-1, .7) {};
			\node[style=none] (timesout) at (-.7, 1){};
			\node[style=small box] (div) at (1, 1) {};
			\node[style=none] (divlab) at (1, 1.5) {$1$};
			\node[style=none] (divin) at (1, .7) {};
			\node[style=none] (divout) at (.7, 1){};

@@

\medskip
If we restrict our attention to separated hypergraphs without isolated nodes, then $\sp$ becomes a right adjoint. The interested reader can consult \Cref{proof:dx} for a proof.

\begin{defi} Let $\mathcal{G}$ be an hypergraph. A node $v\in V_{\mathcal{G}}$ is \emph{isolated} if it does not occur as a letter in either $s_{\mathcal{G}}(e)$ or in
$t_{\mathcal{G}}(e)$ for every $e\in E_{\mathcal{G}}$. $\mathcal{G}$ is \emph{pruned} if it has no isolated nodes. We denote by $\phyp$ and $\pshyp$ the full subcategories of $\hyp$ given by, respectively, pruned and pruned and separated hypergraphs.
\end{defi}

\begin{rem}\label{rem:prun}
	By construction, $\sp(\mathcal{G})$ is pruned for every hypergraph $\mathcal{G}$.
\end{rem}

@@

Finally, we conclude our exam of separated hypergraphs proving a result regarding connected limits. A proof of it is \Cref{proof:seplim}

@@

\smallskip
Since $\mathcal{T}$ must be separated, we have two (non disjoint) cases

\medskip
\begin{minipage}[l]{.65\linewidth}
	\begin{itemize}
		\item  all the hyperedges of $\mathcal{T}$ have sources with length zero; 
		\item all the hyperedges of $\mathcal{T}$ have targets with length zero.
	\end{itemize}
\end{minipage}
\begin{minipage}[r]{.15\linewidth}
\vspace{-.5cm}
\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node[style=medium box] (div) at (5, 1.3) {};
		\node[style=none] (divlab) at (5, 2) {$e$};
		\node[style=none] (divin1) at (4.7, 1.6) {};
		\node[style=none] (divin2) at (4.7, 1) {};
		\node[style=none] (divout) at (5.2, 1.3){};

@@

\smallskip 
\noindent 
\begin{minipage}[l]{.58\linewidth}
\begin{exa}\label{ex:nopush}
	In general $\shyp$ does not have pushouts. For instance consider the solid part of diagram aside. Let us call the two unnamed hypergraphs on the top row $\mathcal{H}$ and $\mathcal{K}$, respectively, while will call the third one $\mathcal{Q}$.

	Suppose that there exists $\mathcal{G}$ and arrows $(p_1, p_2)\colon \mathcal{Q}\to \mathcal{G}$ and $(q_1, q_2)\colon \mathcal{K}\to \mathcal{G}$ which make the square a pushout in $\shyp$. 
	By construction we have
	\[s_{\mathcal{G}}(q_1(e)) = q^\star_2(s_{\mathcal{K}}(e)) = q^\star_2(j^\star_2(v))=p^\star_2(i^\star_2(v)) = p^\star_2(t^\star_{\mathcal{Q}}(e))= t_{\mathcal{G}}(p_2(e))\]
	and so $\mathcal{G}$ cannot be in $\shyp$, yielding a contradiction.
\end{exa}
\end{minipage} \hfill \begin{minipage}[r]{.5\linewidth} \vspace{-2cm}
\xymatrix@R=6pt@C=30pt{\begin{tikzpicture}
		\node[style=node](v) at (0, 0){};
	\node[style=none] at(0, 0.3){$v$};
	\end{tikzpicture}\ar[d]_{(i_1, i_2)} \ar[r]^-{(j_1, j_2)}  &
	\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$j_2(v)$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$w$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, 0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw[style=none] (v) to (in.east); 
		\draw[style=none] (out.east) to (u.west);
	\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[d]^{(q_1, q_2)} \\\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$u$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$i_2(v)$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, -0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw[style=none] (v) to (in.east); 
			\draw[style=none] (out.east) to (u.west);
		\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[r]_-{(p_1, p_2)}&
	\mathcal{G}
	}
\end{minipage}

@@

Our interest for hypergraphs stems from their use as a graphical representation of algebraic terms. As  noted in \Cref{sub:hyp_alg}, an hypergraph can be thought as a colored monoidal signature. Thus labelling corresponds to the operation of taking the slice category over a suitable hypergraph.

@@

\Cref{cor:mono} and \Cref{thm:slice-functors} give an adhesivity result for $\hyp_{\mathcal{G}}$ and a characterisation of its monos.

@@

			Consider an algebraic signature $\Sigma$ and an hypergraph $\mathcal{H}=(E, V, s, t)$. By definition we know that $U_{\hyp}(\mathcal{G}^{\Sigma})$ is the terminal object $1$, so an arrow $\mathcal{H}\rightarrow \mathcal{G}^{\Sigma}$ is determined by a function $h\colon E_\mathcal{H}\to O_\Sigma$  making the two squares on the right commutative (cfr.~\Cref{rem:length}).

@@

As done in \Cref{sub:sub}, we can restrict our attention to two subcategories of labelled hypergraphs.

@@

\begin{prop}
Let $M\colon \lmo \to \hyp$ and $S\colon \shyp \to \hyp$ be the inclusion functors. Let, moreover, $\Delta_{\mathcal{G}}\colon 1\to \hyp$ be the constant functor in $\mathcal{G}$. Then
\begin{enumerate}
	\item $\lmo_{\mathcal{G}}$ is equivalent to $\comma{M}{\Delta_\mathcal{G}}$;
	\item $\shyp_{\mathcal{G}}$ is equivalent to $\comma{S}{\Delta_\mathcal{G}}$; 
\end{enumerate}
\end{prop}

\Cref{lim} allows us to easily characterise regular monos in $\hyp_{\mathcal{G}}$.
\begin{prop}
	A morphism $(f,g)$ between $(h_1, k_1)\colon \mathcal{H}\to \mathcal{G}$ and  $(h_2, k_2)\colon \mathcal{K}\to \mathcal{G}$ is a regular mono in $\lmo_{\mathcal{G}}$ if and only if $(f,g)\colon \mathcal{H}\to \mathcal{K}$ is a regular mono in $\lmo$.
\end{prop}

In turn, by \Cref{thm:slice-functors}, Corollaries \ref{cor:eq}, \ref{cor:pusho} and \Cref{lem:regmon} we get the following result

\begin{thm} Let $\mathcal{G}$ be an hypergraph. Then the following fact hold
	\begin{enumerate}
		\item an arrow $(f,g)$  if and only if is so in $\lmo$; \marginpar{ovvero?}
		\item $\lmo_{\mathcal{G}}$ is quasiadhesive.
	\end{enumerate}
\end{thm}

\begin{rem}\label{rem:tg}
	Let $\Sigma$ be an algebraic signature. Then we know that every hypergraph labelled by $\Sigma$ must be algebraic. By \Cref{rem:elem} this means that an hypergraph $(h,k)\colon \mathcal{H}\to \mathcal{G}^\Sigma$ labelled by $\Sigma$ is left-monogamous if and only if $t_{\mathcal{H}}$ is injective. This is precisely the definition of \emph{term graphs} in \cite{Plu:TGR-ENTCS,CorradiniG05,CastelnovoGM24}. 
\end{rem}

@@

In turn, this allows us to define $\epsilon_{G}\colon V_{\sp(\mathcal{G})}\to V_{\mathcal{G}}$ as the unique one fitting in the diagram below.
